cILR: Taxonomic Enrichment Analysis with Isometric Log-Ratios

### 2021-05-28 tax_agg.tex
Fixed intro section

\begin{abstract}
    \noindent Research in human associated microbiomes primarily involve high-throughput sequencing of samples, resulting in count tables of taxa abundances. Unfortunately, the analysis microbiome data is complex as it is high-dimensional, sparse, and strictly compositional. Various statistical tools have been developed to address these issues, but a very approachable method is to aggregate taxa into pre-defined sets, which is called gene set testing or enrichment analysis in the genomics literature. However, there has been a considerable lack of interest in developing set-based analysis methods tailored specifically for microbiome relative abundance data. Here, we present a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis in the gene set testing literature. Our approach, titled competitive isometric log ratio (cILR), generates enrichment scores per sample as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We also provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance. These results demonstrate how our approach can be used to generate compositionally valid taxonomic aggregation and enrichment.

\end{abstract}

@@

Even though the aforementioned problems are challenging, a very approachable method to address some of them is variable aggregation. Aggregated variables can be less sparse than their constituent elements, and hypothesis testing on a reduced number of variables can alleviate the multiple testing burden. This helps increase power in downstream analyses, as well as interpretability. In practice, microbiome researchers perform variable aggregation to pre-defined taxonomic levels (e.g. phylum, family, order) through element-wise summation of counts. However, this approach contains various downsides: first, it does not allow for comparison of enrichment across sets of different sizes, where larger sets naturally contain more counts; second, aggregating compositional variables using component-wise summations can distort the inter-sample distances before and after aggregation due to the non-linearity of this amalgamation in Aitchison space \cite{egozcue2005}.

Set-based analyses are ubiquitous in the gene expression literature \cite{goeman2007} and various methods have been developed for enrichment testing and scoring. More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. More recent methods, such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate sample level enrichment scores more akin to a transformation. Generated scores can then be used in data visualization, as well as flexible scores that can be utilized in downstream analyses.

Here, we developed a new sample-level set-based analysis method specific to microbiome relative abundance data. We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which compares the expression of genes within the gene set against the rest of the genes. The competitive hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a log-ratio-based aggregation method that addresses the downsides of the naive summation-based method presented above. The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In this manuscript, we present cILR, provide its formulation and discuss some statistical properties. Then, we illustrate the benefits of cILR as both a sample-level significance test for enrichment of taxa sets and as a feature engineering approach for downstream disease prediction and differential abundance using both real data and simulation studies. We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing sample-level enrichment methods in the gene expression literature such as GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

@@

\section*{Discussion}
\subsection*{Set-based analysis can enhance microbiome data analysis}
Gene-set testing (or pathway analysis) is an important tool in the analysis of high-dimensional genomic data sets \cite{khatri2012}. By focusing on a group of variables rather than individuals, this approach reduces the hypothesis testing burden, thereby increasing power as well as replicability. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also increase interpretability.

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum. The process of aggregation involves the element-wise summation of counts of taxa that belong to the same taxonomic group. Even though this allows for a reduction in the number of overall variables (from thousands to only hundreds), there still exist two disadvantages: first, it doesn't preserve inter-sample distances before and after aggregation \cite{egozcue2005}, and second, it doesn't allow for enrichment testing and comparison across sets of different sizes.

As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes. Some software suites, such as \emph{MicrobiomeAnalyst}, have begun to provide an option to perform such analyses with curated taxa sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is based on a hypergeometric test of significant hypotheses and is sensitive to the method of differential abundance used. More recent methods in set-based testing generate enrichment scores at the sample level \cite{hanzelmann2013, frost2020a}. This approach allows for the flexible incorporation of different statistical techniques downstream, but unfortunately is not currently tailored to the compositional nature of microbiome relative abundance data \cite{gloor2017}.

Here, we developed cILR, a novel microbiome-specific method to perform set-based enrichment analysis at the sample level. Our method is based on the isometric log-ratio transformation \cite{egozcue2003}, and measures set enrichment relative to those outside the set, which corresponds to a $Q_1$ null hypothesis in the gene set analysis literature \cite{tian2005}. cILR generates sample-specific enrichment scores, enabling cILR to act as both a significance test for enrichment at the sample level, and as inputs for downstream analyses such as differential abundance testing, predictive modeling, and data visualization.

>>>>

\begin{abstract}
    \noindent Research in human associated microbiomes primarily involve the analysis of taxonomic count tables generated via high-throughput sequencing. It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional. An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets. Set-based analysis is ubiquitous in the genomics literature, and have demonstrable impact in improving interpretability and power of downstream analysis. Current practice in the microbiome field, however, performs aggregation by summing the abundance of respective taxa. This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias. Unfortunately, there is a lack of interest in developing more sophisticated set-based analysis methods specific to microbiome data. Here, we attempt to fill this gap with a new sample-level taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature. Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate under both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios, while additionally providing informative scores that can be inputs to downstream differential abundance and predictive tasks with good performance. These results demonstrate how our approach can enable researchers to generate meaningful aggregation scores for sets of taxa. 
\end{abstract}

@@

Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated sets can be less sparse than their constituent elements, and by focusing on groups rather tahn individuals, this approach can reduce the burden of testing multiple hypotheses, thereby increasing power and reproducibility. Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations. There exists a diverse set of available methods developed in the gene set testing literature. More traditional gene-set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. As such, the most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample and is more akin to a transformation. This strategy allows for the flexible incorporation of different statistical techniques downstream, such as predictive models, as well as for visualization purposes.

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate variables to taxonomic classification levels such as genus, family, or phylum. Standard practice involves summing the abundance of relevant taxa of the same group \cite{mclaren2019}. Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, increase bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}.

As such, there is a need for microbiome researchers to adopt set enrichment methods to perform a more robust set-based analysis that allows for enrichment testing of functionally relevant sets of microbes. Unfortunately, there is a lack of interest in extending existing methods to be more specific to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxa sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is an overrepresentation analysis and therefore similarly sensitive to the method of differential abundance used.

Here, we developed a novel method that allows for generating enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}. We leverage the conception of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which compares the expression of genes within the gene set against the rest of the genes. The competitive hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a log-ratio-based aggregation method that addresses the downsides of the naive summation-based method presented above. The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In the following sections, we provide the formulation of cILR and discuss some statistical properties. Then, we illustrate significance testing at the sample level using cILR and demonstrating type I error and power under different simulation scenarios and real data applications. Then we assessed the informativeness of cILR generated scores, and evaluated how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis. We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are equivalent single sample enrichment analysis methods in the genomics literature. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

@@

\section*{Discussion}
